---
title: "Time Series Final Project"
author: "Steven Cocke"
date: "08/12/19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Read in Data and Plot
This is a time series analysis of the dataset Seatbelts from the datasets package.
In this section I read in the data and do initial plotting: line, ACF and Spectral Density

```{r DriverDeaths, echo=TRUE}
##Load Libraries
library(tswge)
library(datasets)

##Grab dataset
datasets::Seatbelts
?datasets::Seatbelts
#Convert to dataframe
SB <- Seatbelts
class(SB)
df <- data.frame(SB)

#View the structure
str(df)

#View the data
head(df)
```

#### AIC and BIC from AIC5 function. Phi estimation. DriversKilled
In this section I call the aic5 function to determine best model scores for aic and bic for DriversKilled
```{r DriverDeaths1, echo=TRUE}
##AIC ##BIC 
aic5.wge(df$DriversKilled, p=0:15, q=0:10, type = "bic")
aic5.wge(df$DriversKilled, p=0:15, q=0:10)

#Estimate phi
x = est.ar.wge(df$DriversKilled, p = 12)
mean(df$DriversKilled)
```

#### Factor Table to Compare Seasonality - Ensemble Method
In this section I compare quarterly and annual factor tables to the factor table of DriversKilled
to see if there is seasonality present for modeling. Then I attempt to remove the seaonsality and model
and compare this to without seasonality. Standard model vs Ensemble method.

```{r DriverDeaths2, echo=TRUE}
#Factor tables for seasonality
factor.wge(phi = c(0,0,0,0,0,0,0,0,0,0,0,1))
factor.wge(phi = c(0,0,0,1))

#Transform and remove seasonality
y = artrans.wge(df$DriversKilled, phi.tr = c(0,0,0,0,0,0,0,0,0,0,0,1))
y1 = artrans.wge(df$DriversKilled, phi.tr = c(0,0,0,1))

#Plot data with seasonality removed s=12
plotts.sample.wge(y)

#Plot data with seasonality removed s=4
plotts.sample.wge(y1)

#AIC on transformed data s=12
aic5.wge(y, p=0:15, q=0:10, type = "bic")
#AIC on transformed data s=4
aic5.wge(y1, p=0:15, q=0:10, type = "bic")

#Try differences...doesn't help so no d term
xd1.diff = artrans.wge(df$DriversKilled, phi.tr = 1)
xd2.diff = artrans.wge(xd1.diff, phi.tr = 1)
```

#### Estimate and Evaluate with and without Seasonal Components
In this section we estimate and evaluate seasonal and non seasonal models. Compare ASE.
```{r DriverDeaths3, echo=TRUE}
#With seasonal component s=12 AR(1)
x = est.arma.wge(df$DriversKilled, p = 1)
y = fore.aruma.wge(df$DriversKilled, s = 12, phi = x$phi, n.ahead = 15, lastn = TRUE, plot = TRUE, limits = TRUE)
ASE = mean((y$f-df$DriversKilled[178:192])^2)
ASE

#With seasonal component s=4 AR(1)
x = est.arma.wge(df$DriversKilled, p = 12, q=5)
y = fore.aruma.wge(df$DriversKilled, s = 4, phi = x$phi, n.ahead = 15, lastn = TRUE, plot = TRUE, limits = TRUE)
ASE = mean((y$f-df$DriversKilled[178:192])^2)
ASE

#Without seasonal component ARMA(12,5)
x = est.arma.wge(df$DriversKilled, p = 12, q=5)
y = fore.arma.wge(df$DriversKilled, phi = x$phi, theta = x$theta, n.ahead = 15, lastn = TRUE, plot = TRUE, limits = TRUE)
ASE = mean((y$f-df$DriversKilled[178:192])^2)
ASE
```

#### Forecast with and without Seasonal Components
In this section I forecast future 15 months for DriverDeaths.
```{r DriverDeaths4, echo=TRUE}
#With seasonal component s=12 AR(1)
x = est.arma.wge(df$DriversKilled, p = 1)
y = fore.aruma.wge(df$DriversKilled, s = 12, phi = x$phi, n.ahead = 15, lastn = FALSE, plot = TRUE, limits = TRUE)


#With seasonal component s=4 AR(1)
x = est.arma.wge(df$DriversKilled, p = 1)
y = fore.aruma.wge(df$DriversKilled, s = 4, phi = x$phi, n.ahead = 15, lastn = FALSE, plot = TRUE, limits = TRUE)


#Without seasonal component ARMA(12,5)
x = est.arma.wge(df$DriversKilled, p = 12, q=5)
y = fore.arma.wge(df$DriversKilled, phi = x$phi, theta = x$theta, n.ahead = 15, lastn = FALSE, plot = TRUE, limits = TRUE)
```

#### VAR Analysis - add in kms driven attribute to the analysis
In this sections is the initial steps for the VAR modeling. Separate out test and training and rename columns.
```{r DriverDeaths5, echo=TRUE}
##Plot ccf for Kms and DriversKilled
ccf(df$kms, df$DriversKilled, ylim = c(-1,1))

##Take all except last 15 month recordings - we will use this for testing
df_kms = df$kms[1:177]
df_deaths = df$DriversKilled[1:177]
all_data = data.frame(cbind(df$kms, df$DriversKilled))
names(all_data) <- c("kms","DriversKilled")
```

#### VAR Analysis - BIC for kms driven and DriversKilled
In this section I find the BIC for kms and DriversKilled
```{r DriverDeaths6, echo=TRUE}
##kms
p_df_kms = aic5.wge(df_kms, p=0:15, q=0:10, type = "bic")
p_df_kms

##driver deaths
p_df_deaths = aic5.wge(df_deaths, p=0:15, q=0:10, type = "bic")
p_df_deaths
```

#### VAR Analysis - Estimation and Forecast Univariate Regression
In this section I estimate and forecast for univariate regression models separately for kms driven and
Driver Deaths. 
```{r DriverDeaths7, echo=TRUE}
#estimate and forecast kms Univariate Regression
kms.est = est.ar.wge(df_kms, p = 15)
fore.arma.wge(df_kms, phi = kms.est$phi, n.ahead = 15, lastn = FALSE, limits = TRUE)

#estimate and forecast kms Univariate Regression
deaths.est = est.ar.wge(df_deaths, p = 12)
fore.arma.wge(df_deaths, phi = deaths.est$phi, n.ahead = 15, lastn = FALSE, limits = TRUE)
```

#### VAR Analysis - Estimation and Evaluate Multivariate
In this section I estimate and forecast for multivariate using VARSelect and combine the attributes into the model.
We check the ASE and also plot to see performance.
```{r DriverDeaths8, echo=TRUE}
#VAR SELECT
X = cbind(df_kms, df_deaths)
VARselect(X, lag.max = 16, type = "const", season = NULL, exogen = NULL)

#Fit
lsfit = VAR(X, p=14, type = 'const')

#Predictions 15 ahead
preds = predict(lsfit, n.ahead = 15)
preds

#ASE kms
preds$fcst$df_kms[1:15,1]
ASE = mean((preds$fcst$df_kms[1:15,1]-df$kms[178:192])^2)
ASE

#ASE deaths
preds$fcst$df_deaths[1:15,1]
ASE = mean((preds$fcst$df_deaths[1:15,1]-df$DriversKilled[178:192])^2)
ASE

#plot Deaths
plot(seq(1,192,1), df$DriversKilled, type = "b")
points(seq(178, 192, 1), preds$fcst$df_deaths[1:15,1], type = "b", pch = 15)

#plot kms
plot(seq(1,192,1), df$kms, type = "b")
points(seq(178, 192, 1), preds$fcst$df_kms[1:15,1], type = "b", pch = 15)

#fanchart
fanchart(preds)
```

#### VAR Analysis - Forecast Next 15 months Multivariate
In this section I forecast the next 15 months for future data points.
```{r DriverDeaths9, echo=TRUE}
##forecast using all the data; have to combine data for plotting
X = cbind(df$kms, df$DriversKilled)
VARselect(X, lag.max = 16, type = "const", season = NULL, exogen = NULL)
lsfit = VAR(X, p=14, type = 'const')
preds = predict(lsfit, n.ahead = 15)
preds

#predictions
prediction = data.frame(cbind(preds$fcst$y1[,"fcst"],preds$fcst$y2[,"fcst"]))
names(prediction) <- c("kms","DriversKilled")
forecast = rbind(all_data, prediction)

#forecasts for deaths
plot(seq(1,207,1), forecast$DriversKilled, type = "b")
points(seq(193, 207, 1), forecast$DriversKilled[193:207], type = "b", pch = 15)

#forecasts for kms
plot(seq(1,207,1), forecast$kms, type = "b")
points(seq(193, 207, 1), forecast$kms[193:207], type = "b", pch = 15)
```

#### Neural Network Analysis - Time as the Only Regressor
In this section I create a Neural Network model using only time as the regressor for DriverDeaths
```{r DriverDeaths10, echo=TRUE}
library(nnfor)
df_ts = ts(df$DriversKilled[1:177], start = c(1969,1), frequency = 12)
df_ts_test = ts(df$DriversKilled[178:192], start = c(1983,10), frequency = 12)

#only time as regressor
fit.mlp = mlp(df_ts, reps = 100)
fit.mlp

#plot the NN
plot(fit.mlp)

#evaluate the NN with ASE on the 15 last months held out for testing
f = forecast(fit.mlp, h=15)

#plot the results
plot(df$DriversKilled[178:192], type = "l")
lines(seq(1,15),f$mean, col = "blue")

#ASE
ASE = mean((df_ts_test-f$mean)^2)
ASE
```

#### Neural Network Analysis - Additional Regressors
In this section I create a Neural Network model adding in PetrolPrice, and number of Drivers
```{r DriverDeaths11, echo=TRUE}
#with additional regressors
df_tsx = data.frame(kms = ts(df$kms), petrol = ts(df$PetrolPrice), drivers = ts(df$drivers))

#fit
fit2 = mlp(df_ts, xreg = df_tsx, reps = 100)

#evaluate
f2 = forecast(fit2, h = 15, xreg = df_tsx)

#plot the results
plot(df$DriversKilled[178:192], type = "l", ylim = c(60, 160))
lines(seq(1,15), f2$mean, col= "blue")

#ASE
ASE = mean((df$DriversKilled[178:192]-f2$mean)^2)
ASE
```